Untitled

### 2022-06-21 Manuscript/Regression_trees.Rmd
edits

---
title: \singlespacing \vspace{-1.6cm} \LARGE Trade-offs among plant reproductive traits determine interactions with floral visitors

#author: 
#  - Jose B. Lanuza^1,2^*
#  - Romina Rader^1^
#  - Jamie Stavert^3^
#  - Liam K. Kendall^4^
#  - Manu E. Saunders^1^
#  - Ignasi Bartomeus^2^

# may use lua filters for affiliations etc: see https://stackoverflow.com/a/52919256

# date: '`r Sys.Date()`'

output:
  bookdown::pdf_document2:
    fig_caption: yes
    fig_width: 7
    fig_height: 7
    df_print: kable
    keep_tex: yes
    number_sections: no
    extra_dependencies: ["float"]
    toc: no
    fig_crop: no
    includes:
      in_header: header.tex

documentclass: article
# classoption: twocolumn, landscape, etc
papersize: a4
pagestyle: plain  # plain, empty, headings
geometry: margin=1in
linestretch: 2
fontfamily: mathpazo
fontsize: 12pt

bibliography: 
  - references.bib
  - knitcitations.bib
# zotero: "My shared Zotero group"
csl: ecology.csl #ecology letters
notes-after-punctuation: false
link-citations: yes
linkcolor: RoyalBlue
urlcolor: RoyalBlue
links-as-notes: false
subparagraph: yes

header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
- \usepackage{float}
- \floatplacement{figure}{H}
- \usepackage[skip=3pt]{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{setspace}
- \usepackage{titlesec}
- \titlespacing{\title}{0pt}{\parskip}{-\parskip}
- \usepackage[style=nature]{biblatex}
- \usepackage[labelformat = empty]{caption}
- \usepackage{pdflscape}
#- \DeclareUnicodeCharacter{0301}{*************************************}

---

\vspace{-1.4cm}

\singlespacing

\textbf{Jose B. Lanuza$^{1,2}$} <barragansljose@gmail.com>, \textbf{Romina Rader$^{1}$} <rrader@une.edu.au>, \textbf{Jamie Stavert$^{3}$} <jamie.stavert@gmail.com>, \textbf{Liam K. Kendall$^{4}$} <liam.k.kendall@gmail.com>, \textbf{Manu E. Saunders$^{1}$} <Manu.Saunders@une.edu.au> \textbf{and Ignasi Bartomeus$^{2}$} <nacho.bartomeus@gmail.com>

\small

^1^ School of Environmental and Rural Science, University of New England, Armidale, New South Wales 2350, Australia. ^2^ Estación Biológica de Doñana (EBD-CSIC), E-41092 Seville, Spain. ^3^ Department of Conservation | Te Papa Atawhai, Auckland, New Zealand. ^4^ Centre for Environmental and Climate Science, Lund University, Sölvegatan 37, S-223 62 Lund, Sweden.

\doublespacing
\normalsize

\textbf{Statement of authorship:} JBL, RR and IB designed the study. JBL collated the data and conducted analysis with guidance of JS, LKK and IB. JBL wrote the manuscript with contributions of all authors.

\textbf{Data accessibility statement:} All data and code used to conduct this study will be deposited in the public repository Dryad upon acceptance.

\textbf{Short title:} Plant reproductive trade-offs.

\textbf{Keywords:} life-history strategies | plant reproductive trade-offs | plant-pollinator interactions.

\textbf{Type of article:} Letter.

\textbf{Number of words:} Abstract  (146 words), main text excluding references, acknowledgements and captions (4983 words).

\textbf{Number of references:} 75

\textbf{Number of figures and tables:} 4 figures and 1 table.

\textbf{Corresponding author:} Jose B. Lanuza | +34 616375981| <barragansljose@gmail.com>

\newpage

# ABSTRACT

Plant life-history strategies are constrained by cost-benefit trade-offs that determine plant form and function. However, despite recent advances in the understanding of plant trait variation, little is known about plant reproductive trade-offs and how these constrain life-history strategies and shape interactions with floral visitors. Here, we investigate plant reproductive trade-offs and how these drive interactions with floral visitors using a dataset of 16 reproductive traits for 1,506 plant species. We found that over half of all plant reproductive trait variation was explained by two independent axes. Specifically, the first axis indicated the presence of a trade-off between flower number and flower size, while the second axis indicated a pollinator dependency trade-off. In addition, plant reproductive trade-offs determined important differences in the interaction level among floral visitor guilds. Our study shows the main reproductive trade-offs of flowering plants and their relevance to understand plant-pollinator interactions in a global context.

\newpage

\doublespacing
\vspace{5mm}
\normalsize

```{r setup, include=FALSE, cache=FALSE, message = FALSE}

library("knitr")

### Chunk options: see http://yihui.name/knitr/options/ ###
## Text results
opts_chunk$set(echo = FALSE, warning = TRUE, message = TRUE)

## Code decoration
opts_chunk$set(tidy = TRUE, comment = NA, highlight = TRUE)

## Cache
# opts_chunk$set(cache = 2, cache.path = "output/cache/")

## Plots
opts_chunk$set(fig.path = "output/figures/")

```{r knitcitations, cache = FALSE}
library(knitcitations)
cleanbib()   
cite_options(citation_format = "pandoc")

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

# INTRODUCTION

Flowering plants have an astonishing diversity of floral structures [@barrett2002; @schiestl2013] that shape plant-pollinator associations [@fenster2004; @dellinger2020]. However, not all reproductive trait combinations are possible due to evolutionary and ecological constraints [REF]. Despite the recent advances in the theoretical understanding of the major floral trade-offs [e.g., floral economic spectrum; @roddy2021], we lack empirical evidence that helps to elucidate the spectrum of reproductive trait variation at a global scale as done recently for other vegetative and physiological traits [REF].

Although there is an increasing number of macroecological studies that investigate plant reproductive traits [@baude2016; @munoz2016; @grossenbacher2017; @moeller2017], we still have poor understanding of how reproductive traits drive interactions with pollinators at large scales [@sargent2008; @rech2016; @salguero2016; @ruger2018]. Because the species' morphology can determine the species' functional role, by linking the plant's position in the reproductive trait-space with the different pollinator groups [as done similarly in @dehling2016], we could help to improve our understanding of plant-pollinator associations. Further, there is increasing interest in understanding drivers of plant-pollinator interactions using trait-based approaches [@fenster2004; @rosas2014] and trait-matching analyses [@stang2009; @bartomeus2016]. However, plant reproductive traits have been overlooked for most pollination systems [@dellinger2020] and remains unclear how specific plant reproductive biology traits (e.g., mating or compatibility system) influence plant-pollinator associations [@tur2013; @devaux2014].

Species can optimise their fitness through various life-history traits, yet trade-offs among those traits constrain the range of potential strategies that a species can use. With the recent availability of large trait databases [e.g., TRY @kattge2011; and COMPADRE @salguero2015], plant ecological strategies are being increasingly examined, and are facilitating the identification of global patterns and constraints in plant form and function [@salguero2016; @diaz2016; @bruelheide2018; @carmona2021]. However, most studies have focused on vegetative traits such as leaf [@wright2004], wood [@chave2009], or root [@laughlin2021] trade-offs with little or no attention given to reproductive traits [@evojtko2020; @roddy2021] which are critical to plant life strategies that shape interactions with pollinators and ultimately determine plant reproductive success. For instance, short lived versus perennial species tend to have low versus high levels of outcrossing, respectively, [@barrett2003; @moeller2017] and outcrossing levels are positively correlated with flower size [@goodwillie2010]. In addition, the presence of costly rewards (e.g., pollen or nectar) and showy flowers or floral displays can only be understood through consideration of plant species’ reliance upon animal pollination (pollinator dependence) and its role in attracting pollinators [@ollerton2011; @rodger2021]. However, it is still unknown to what extent these different reproductive strategies determine plant-pollinator interactions.

Several studies have identified links between plant traits and plant-pollinator network properties [@bartomeus2013; @olito2015; @rowe2020]. Moreover, plant traits can define species’ network roles [e.g., specialists vs generalists\; @lazaro2013; @tur2013]. For example, plant species that occupy reproductive trait space extremes are more likely to exhibit higher levels of specialisation and be more reliant on the trait-matching with pollinators [@junker2013; @coux2016]. Morphological matching between plant and floral visitors often determines plant-pollinator interactions, and can thus strongly influence interaction network structure [@stang2009; @ibanez2012]. Importantly, the combination of traits has shown to increase the predictive power of the network interactions [@eklof2013] and by considering the position in the multidimensional trait space we can determine species’ roles within plant-pollinator interaction networks [@dehling2016]. Nonetheless, we know little if those patterns generally studied at the community level are representative of wider macroecological scales.

Here, we aim to explore the potential trade-offs among plant reproductive traits and their association with the different floral visitor guilds. First, we identify the major axes of reproductive trait variation and trade-offs that determine plant form and function. Second, we investigate the association between the plant species’ position in trait-space and the different floral visitor guilds with the help of qualitative (presence-absence of interaction) and quantitative (number of visits) information. Finally, we investigate how both the main axes of trait variation, and individual traits, influence plant species’ roles within networks using a set of complementary interaction network metrics (i.e., number of visits, normalized degree and specialization).

# MATERIALS AND METHODS

**Plant-pollinator network studies**

We selected 28 studies from 18 different countries that constituted a total of 64 plant-pollinator networks (see Table S1 and Fig. S1). These studies recorded plant-pollinator interactions in natural systems and were selected so that we had broad geographical representation. Although these studies differ in sampling effort and methodology, all studies provided information about plant-pollinator interactions (weighted and non-weighted), which we used to build a database of plant species that are likely to be animal pollinated. Many of these networks are freely available either as published studies [@olesen2007; @fortuna2010; @carvalheiro2014] or available in online archives [e.g., The Web of Life, @fortuna2010; and Mangal, @poisot2016]. In total, our network dataset constituted 60 weighted (number of visits) and 4 unweighted (presence-absence of the interaction) networks, each sampled at a unique location and year, as well as eight meta-webs where interactions were pooled across several locations and multiple years.

**Taxonomy of plants and pollinators**

All species names, genera, families and orders were retrieved and standardized from the taxonomy data sources NCBI (https://www.ncbi.nlm.nih.gov/taxonomy) for plants and ITIS (https://www.itis.gov/) for pollinators, using the R package _taxize_ [@chamberlain2020]. We filled the ‘not found’ searches manually using http://www.theplantlist.org/ and http://www.mobot.org/ for plants and http://www.catalogueoflife.org/ for floral visitors.

**Plant traits**

We selected a total of 19 different functional traits that included both reproductive and vegetative traits (see Table 1 and Supplementary Information). From these, 16 were reproductive traits (13 floral and 3 reproductive biology traits) and were selected based on their relevance to plant reproduction and data availability. The 3 remaining traits, were vegetative traits that are commonly used to characterize the global spectrum of plant form and function and represent the fast-slow continuum of trait variation (e.g., short-lived versus perennial species). For each plant species, we undertook an extensive literature and online search across a wide range of resources (plant databases, online floras, books, journals and images). From a total of 30,120 cells (20 columns × 1,506 species) we were able to fill 24,341 cells (80.8% of the dataset, see Fig. S2 for missing values information for each trait).

**Phylogenetic Distance**

We calculated the phylogenetic distance between different plant species using the function *get_tree* from the package _rtrees_ (https://github.com/daijiang/rtrees), which downloads phylogenetic distances from the extended R implementation of the Open Tree of Life [@smith2018; @jin2019].

**Data Imputation**

Trait missing values were imputed with the function _missForest_ [@stekhoven2012] which allows imputation of data sets with continuous and categorical variables. We accounted for the phylogenetic distance among species on the imputation process by including the eigenvectors of a principal component analysis of the phylogenetic distance (PCoA) which has been shown to improve the performance of _missForest_ [@penone2014]. To extract the eigenvectors, we used the function _PVRdecomp_ from the package _PVR_ [@santos2018] based on a previous conceptual framework that considers phylogenetic eigenvectors [@diniz-filho2012]. We conducted two different imputations, one for the full set of species (1,506 species, 5.79% of missing values) excluding nectar and pollen traits because of the high percentage of missing values (Fig. S2) and a second one for the subset of species with data for pollen per flower and microliters of nectar (755 species, 8.01% of missing values).

**Plant strategies**

We explored the trade-offs between the different quantitative plant traits with a phylogenetically informed Principal Component Analysis (pPCA). We did not include the quantitative variables of flower length and inflorescence width because they were highly and moderately correlated to flower width respectively (Pearson’s correlation = 0.72, _P_ < 0.01 and Pearson’s correlation = 0.36, _P_ < 0.01), and thus we avoided overemphasizing flower size on the spectrum of trait variation. Although qualitative traits were not included in the dimensionality reduction analysis, we also investigated the association of the different qualitative traits with the main axes of trait variation. Prior to the analyses, we excluded outliers and standardized the data. Due to the high sensitivity of dimensionality reduction to outliers, we excluded values within the 2.5th–97.5th percentile range [@legendre2012], and thus our final dataset had 1,236 species. Then, we log transformed the variables to reduce the influence of outliers and z-transformed (X= 0, SD=1) so that all variables were within the same numerical range. We performed the pPCA using the function _phyl.pca_ from the package _phytools_ [@revell2012] with the method lambda ($\lambda$) that calculates the phylogenetic correlation between 0 (phylogenetic independence) and 1 (shared evolutionary history) and we implemented the mode covariance because values for each variables were on the same scale following transformation [@abdi2010]. Moreover, to corroborate that our imputation of missing values did not affect our results, we conducted a pPCA on the full dataset without missing values (Fig. S3). We found little difference between the explained variance with the imputed dataset (51.08%) and the dataset without missing values (52.87%). In addition, the loadings on each principal component had a similar contribution and correlation patterns, with the exception of plant height which showed slight variations between the imputed and non-imputed dataset. Finally, we conducted an additional phylogenetic informed principal component analysis for the subset of species with pollen and nectar quantity. For this, we included all quantitative traits considered in the main pPCA plus pollen grains and microlitres of nectar per flower.

**Phylogenetic signal**

We calculated the phylogenetic signal of the different quantitative traits on the imputed dataset with the full set of species (N = 1,506) with the package _phytools_ [@revell2012] and we used Pagel’s $\lambda$ as a measurement of the phylogenetic signal. However, for pollen and nectar traits, phylogenetic signal was calculated only on the subset of species that had quantitative information for these traits (N = 755).

**Network analyses**

Analyses were conducted on the subset of 60 weighted networks sampled in a unique flowering season and site, which included 556 plant and 1,126 pollinator species. These networks were analysed in their qualitative and quantitative form. First, we analysed the binary version of these weighted networks with presence-absence information that assumes equal weight across interactions. Second, we analysed the untransformed weighted networks with the number of visits of floral visitors to individual flowers that account for the intensity of the interaction. Although floral visitors are not always pollinators and number of visits does not consider each pollinator species efficiency [@ballantyne2015], the number of visits can provide valuable information of the contribution of floral visitors to pollination [@vazquez2005; @vazquez2012]. In total, our network dataset (excluding meta-webs and non-weighted networks) included 2,256 interactions of bees with plants, 1,768 non-syrphid-Diptera interactions, 845 syrphids interactions, 437 Lepidoptera interactions, 432 Coleoptera interactions and 362 non-bee-Hymenoptera interactions. Sampling methods varied across networks but this was accounted for in analyses by considering them in the random effects of the modelling process. All analyses were conducted in R version _4.0.3_.

**Visitation patterns**

We used Bayesian modelling (see below for details) to explore the effect of floral visitor groups and the main axes of trait variation (pPCA with imputed dataset) on both qualitative and quantitative floral interactions per plant species. For this, we divided floral visitors into six main guilds that differ in life form, behaviour and are likely to play a similar ecological role: (i) bees (Hymenoptera-Anthophila), (ii) non-bee-Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Diptera-Syrphidae), (iv) non-syrphid-Diptera (Diptera-non-Syrphidae), (v) Lepidoptera and (vi) Coleoptera. Moreover, because the guild of bees was the most represented group with 2,256 records and had the highest frequency of visits, we also explored presence-absence of the interaction and number of visits of the main bee families (Andrenidae, Apidae, Colletidae, Halictidae and Megachilidae) on the trait space. In addition, we found that _Apis mellifera_ was the floral visitor with the largest proportion of records counted (7.55% of the total). This finding is consistent with previous research showing that _A. mellifera_ was the most frequent floral visitor in a similar dataset of 80 plant-pollinator networks in natural ecosystems [@hung2018]. Hence, to control for the effect of _A. mellifera_ on the observed visitation patterns of bees, we conducted an analogous analysis with presence-absence of the interaction and number of visits excluding _A. mellifera_. We found that _A. mellifera_, was partly driving some of the observed trends on PC1 (Fig. S4). However, we did not detect major differences on PC2 and PC3.

We implemented Bayesian generalized linear mixed models using the R package _brms_ [@burkner2017]. We modelled presence-absence of observed interactions and number of visits as a function of the main axes of plant trait variation and their interactions with floral visitor guilds (e.g., number of visits ~ PC1 x FGs + PC2 x FGs + PC3 x FGs). Because we were interested in possible differences in the visitation patterns among floral visitors groups to plants with different strategies, we included interactions between the main axes of trait variation (PC1, PC2 and PC3) and the different floral visitor guilds. We added a nested random effect of networks nested within the study system to capture the variation in networks among studies and within networks. Moreover, we included the phylogenetic covariance matrix as a random factor due to the possible shared evolutionary histories of species and therefore lack of independence across them. We specified for presence-absence of interaction and number of visits a Bernoulli and a zero inflated negative binomial distribution, respectively. The models were run with non or very weakly informative informative priors from the brm function so they have neglible influence on the results [@burkner2017], 3,000 iterations and with previous 1,000 warm up iterations. We set delta ($\Delta$) to 0.99 to avoid divergent transitions and visualized the posterior predictive checks with the function *pp_check* using the _bayesplot_ package [@gabry2019].

**Plant species network roles**

We investigated whether different quantitative traits determined plant species network roles using Bayesian modelling and regression trees. For this, we selected simple and complementary species-level network metrics commonly applied in bipartite network studies [@dormann2008] with a straightforward ecological interpretation relevant to our research goals. The different plant species-level metrics were: (i) sum of visits per plant species; (ii) normalized degree, calculated as the number of links per plant species divided by the total possible number of partners; and (iii) specialization (d’) [@bluthgen2006], which measures the deviation of an expected random choice of the available interaction partners and ranges between 0 (maximum generalization) and 1 (maximum specialization). Normalized degree and specialization were calculated with the _specieslevel_ function from the R package _bipartite_ [@dormann2008].

First, we modelled the distinct plant species metrics (sum of visits, normalized degree and plant specialization) as a function of the three main axes of trait variation (plant species level metric ~ PC1 + PC2 + PC3). For each response variable (i.e., each plant species level metric), we used different distribution families (zero inflated negative binomial for the sum of visits, weibull for normalized degree and zero one inflated beta for specialization). Finally, we used the same random factors, model settings and conducted the same posterior predictive checks for each model as detailed above in the ‘visitation patterns’ section.

Second, to better understand complex trait relationships, we used regression trees. Regression trees are recursive algorithms which can detect complex relationships among predictors and allow identification of the relevance of specific trait combinations on explaining species roles within the network of interaction. We focused exclusively on quantitative traits because almost all categorical traits were statistically associated with the first two axes of trait variation (Table S2). We conducted this analysis using the _rpart_ package [@therneau2015] with method _‘anova’_ with a minimum of 50 observations per terminal node and we used the _rpart.plot_ package [@milborrow2015] to plot the regression trees. We considered the species level indices as response variables (number of visits, normalized degree and specialization) and we performed one regression tree per metric using the different quantitative traits as predictors. We calculated two regression trees per plant species-level metric, one for the full set of species and another for the subset of species for which we had pollen and nectar traits. We focused on regression trees that included floral rewards because they consistently showed pollen and nectar traits as being the best for explaining the different species-level metrics (Fig. S5).

# RESULTS

**Plant strategies**

The phylogenetically informed principal component analysis captured by the first two and three axes 51.8% and 70.97% of trait variation, respectively (Fig. 1 and Fig. S6) and had a phylogenetic correlation ($\lambda$) of 0.76. The first principal component (PC1) represented 26.72% of the trait variation and indicated a trade-off between flower number and flower size. We refer to this axis as the ‘flower number - flower size trade-off’, as already described in previous studies [@sargent2007; @kettle2011]. Hence, one end of the spectrum comprised species with high investment in flower number and plant height but small flower size, short style length and low ovule number. The other end of this spectrum comprised species that were short in height and invested in large flowers, long styles, many ovules, but few flowers. The main contributing traits to PC1 were plant height, flower number, ovule number and flower size (loadings > |0.5|; Table S3) but style length also contributed moderately to PC1 (loading = -0.33). The second principal component (PC2) represented 25.05% of the trait variation and indicated a trade-off between low and high pollinator dependence. We refer to this axis as the ‘pollinator dependence trade-off’. The main driver of trait variation on PC2 was autonomous selfing (loading = 0.85) but the other traits (except ovule number) also made moderate contributions (loadings from 0.27 to 0.4; Table S3). We found that high pollinator dependence was associated with larger and a higher number of flowers, greater plant height and longer styles. In contrast, species with high levels of autonomous selfing tended to have fewer and smaller flowers, had shorter styles and were shorter in height. Further, PC3 explained a considerable amount of trait variability (19.17%) and the main contributors to this axis were style length (loading = -0.66) and the degree of autonomous selfing (loading = -0.51). The remaining traits, apart from ovule number, were moderately correlated to changes on PC3 (loadings from -0.23 to -0.46; Table S3). Thus, because style length was correlated with all traits on PC3 and was the main driver of trait variation, we refer to this axis as the ‘style length trade-off’. Further, the pPCA with the subset of species that had nectar and pollen quantity data showed that nectar quantity (microlitres of nectar per flower) and pollen grains per flower were positively associated with flower size, style length and ovule number but negatively associated with flower number (PC1, 26.82%; Fig. S7). This pPCA explained similar variance with the first two principal components (45.52%) and similar associations of traits despite some variability in the loadings (Table S4).

We found that most categorical traits were statistically associated with the first two axes of trait variation (Fig. 2 and Table S2). Flower symmetry, which was only associated with PC2 (Sum of squares = 8.51, F-value = 14.72, _P_ < 0.01 ), and nectar provision, which was independent of PC1 and PC2 (PC1: Sum of squares = 0.37, F-value = 0.29 , _P_ = 0.59; PC2:  Sum of squares = 0.83, F-value = 1.43, _P_ = 0.23) showed lack of statistical association. In addition, we found (with a Tukey test) statistical differences between the different levels of categorical traits in the trait space (Fig. S8). Regarding self compatibility, we found larger differences on PC2 (i.e., species with unisexual flowers that were self incompatible were statistically differentiated from species with partial or full self compatibility; Fig. S8a and Fig. S8b). Life forms differed statistically across both axes of trait variation and followed a gradient of larger life forms (trees and shrubs) with higher pollinator dependence to smaller ones (herbs) with lower pollinator dependence (Fig. S8c and Fig. S8d). Consequently, lifespan also followed this gradient but perennial and short lived species only differed statistically on PC2 (Fig. S8e and Fig. S8f). Species with unisexual flowers (monoecious and dioecious) were clustered on both extremes of the first two principal components and had the highest pollinator dependence and highest number of flowers (Fig. S8g and Fig. S8h). Moreover, we found that the campanulate and capitulum flower shapes were differentiated from tube, papilionaceous, open and brush shapes in the trait space. The former morphologies had larger flowers and greater pollinator dependence, while the latter had higher flower number and greater autonomous selfing (Fig. S8i and Fig. S8j). Regarding flower symmetry, zygomorphic flowers were associated with lower levels of pollinator dependence, whereas actinomorphic flowers had higher levels of pollinator dependence (Fig. S8k and Fig. S8l).

**Phylogenetic signal**

We found a strong phylogenetic signal (*P* < 0.01) in all quantitative traits (Table S5). The traits that showed the highest phylogenetic signal were ovule number ($\lambda$ = 1), pollen grains per flower ($\lambda$ = 1) and plant height ($\lambda$ = 0.96), followed by flower length ($\lambda$ = 0.75), flower width ($\lambda$ = 0.73), number of flowers per plant ($\lambda$ = 0.69) and nectar concentration ($\lambda$ = 0.65). The traits that showed a moderate phylogenetic signal were inflorescence width ($\lambda$ = 0.57), style length ($\lambda$ = 0.49) and autonomous selfing ($\lambda$ = 0.34). Finally, microliters of nectar per flower showed the lowest phylogenetic signal of all traits ($\lambda$ = 0.14).

**Visitation patterns**

The main axes of trait variation explained partly presence-absence of interaction partners (conditional $R^{2}$ = 0.26; marginal $R^{2}$ = 0.20) but little of the overall number of visits (conditional $R^{2}$ = 0.31; marginal $R^{2}$ = 0.06). However, we found relevant differences across the different floral visitor guilds on both presence-absence of interactions and number of visits (Fig. 3). We found that plants with high flower number and small flowers had higher interaction partners of Coleoptera, non-bee-Hymenoptera and all Diptera guilds but plants with low flower number but large flowers had higher interaction partners of bees and Lepidoptera guilds (flower number - flower size trade-off, PC1; Fig. 3a). However, plant species with high flower number but small flowers had higher number of visits of bees and syrphids guilds (PC1; Fig. 3d). Remarkably, all plant species with higher pollinator dependence had higher number of interacting partners and number of visits for all floral visitor guilds (PC2; Fig. 3b and Fig. 3e). Finally, plant species with short styles and low selfing had higher interaction partners of all guilds but bees that interacted clearly more with plant species with long styles and high selfing (style length trade-off; Fig. 3c). However, for number of visits, we found that plants with long styles and high selfing interacted more frequently with Lepidoptera and non-bee-Hymenoptera guilds (Fig. 3f).

The additional model for both presence-absence of interaction (marginal $R^{2}$ = 0.29; conditional $R^{2}$ = 0.19) and number of visits (marginal $R^{2}$ = 0.30; conditional $R^{2}$ = 0.03) for the most represented families of bees showed that the family Apidae was the main driver of the observed patterns. The contrasting differences between presence-absence of interaction and number of visits for bees on PC1 (Fig. 3a and Fig. 3d) were driven by the family Andrenidae that had higher number of interacting partners but lower number of visits on plant species with low number of large flowers (Fig. S9).

**Plant species network roles**

The variance of the different plant species-level network metrics was poorly explained by the three main axes of trait variation (Fig. S10; number of visits ~ PCs, conditional $R^{2}$ = 0.11, marginal $R^{2}$ = 0.02; normalized degree ~ PCs, conditional $R^{2}$ = 0.24, marginal $R^{2}$ = 0.02; and, specialization ~ PCs, conditional $R^{2}$ = 0.37, marginal$R^{2}$ = 0.03). Overall, the most notable trends were found on PC1 and PC3 for number of visits and specialization. On the flower number - flower size trade-off (PC1), number of visits was higher for plant species with more flowers but was lower for plant species with larger flowers (Fig. S10a). On PC1, specialization showed the opposite trend (Fig. S10g). On the style length trade-off (PC3), number of visits was lower for plants with shorter styles and lower autonomous selfing and higher for species with longer styles and higher autonomous selfing (Fig. S10c). Again, specialization showed the opposite trend to number of visits (Fig. S10i).

When we further investigated the combination of traits that drive plant network roles, we found that the regression tree for number of visits was best explained by plant height, nectar concentration and style length (Fig. 4a). Specifically, species taller than 3.9m had the highest number of visits, while species that were shorter than 3.9m and had a nectar concentration lower than 16% had the lowest number of visits. Normalized degree was best explained by nectar concentration, pollen grains per flower, plant height, flower width and autonomous selfing (Fig. 4b). Species with a nectar concentration over 49% had the highest levels of normalized degree, whereas species with nectar concentration lower than 49%, more than 21,000 pollen grains per flower and height less than 0.78m had the lowest normalized degree. Finally, specialization was best explained by plant height, ovule number, pollen grains per flower and autonomous selfing (Fig. 4c). Overall, plant species with the highest specialization were shorter than 1.3m, had more than 14,000 pollen grains per flower and autonomously self-pollinated less than 11% of their fruits. In contrast, species taller or equal than 5.1m and with lower than 14 ovules per flower had the lowest specialization values.

# DISCUSSION

This study demonstrates that plant species exhibit clear trade-offs among their vegetative and reproductive traits and that these trade-offs determine interactions with floral visitors. These trade-offs are differentiated along three axes of trait variation: (i) flower number - flower size, (ii) pollinator dependence and (iii) style length. These reproductive trade-offs helped partly explain the presence of floral visitor interactions, but not their number of visits. However, floral visitor guilds formed distinct relationships with the main axes of trait variation. Moreover, we found that the plant species network roles were best explained by plant size and floral reward related traits.

Over half of all plant trait variation was captured by the flower number - flower size and pollinator dependence trade-offs. Trait variation on these two axes was associated with the ‘fast-slow continuum’ in plant [@salguero2016] and animal [@healy2019] life-history strategies, as indicated by the different floral and reproductive biology traits associated with plant height, life form and lifespan. The ‘slow’ part of this continuum (i.e., tall trees and shrubs) included plant species with many flowers, few ovules, higher pollinator dependence, frequent occurrence of self-incompatibility and more complex breeding systems (e.g., monoecious and dioecious species). In contrast, plant species that employed the ‘fast’ strategy (i.e., short herbs), had fewer flowers, more ovules, frequent occurrence of self-compatibility and lower pollinator dependence. Further, on the first two axes of trait variation, we found additional support for the previously described positive association between higher outcrossing rate and larger floral display [@goodwillie2010]. The positive correlation between larger floral display and higher pollinator dependence in our dataset further confirmed this trend (Fig. S11).

Despite the low predictive power of the main trait variation axes for broad-level interaction patterns (number of interaction partners and number of visits), we found changes in the interaction patterns among and within floral visitor guilds across these axes that suggest plant life-history strategies influence plant-pollinator interactions. For example, all floral visitor guilds visited plant species with higher pollinator dependence more frequently, and high pollinator dependence was associated with large floral displays and greater pollen quantities (Fig. 1 and Fig. S6). This trend is consistent with previous studies that show plant species with higher reproductive investment tend to be visited by pollinators more frequently [@hegland2005; @lazaro2013; @kaiser2014]. In regard to the flower number - flower size and style length trade-offs, different pollinator guilds showed contrasting visitation patterns across the continuum of trait variation, which could be associated with different pollination syndromes at a macroecological scale. For instance, bees and syrphid flies were clearly associated with opposing life-strategies on PC1 and PC3 (Fig. 3) suggesting possible niche partitioning [@palmer2003; @phillips2020] between these two guilds. However, despite floral rewards not being included in the main analysis because there was insufficient data available, floral reward related traits were among the best at characterising species network roles (Fig. 4). More detailed exploration of reproductive trade-offs in conjunction with floral rewards is needed to help elucidate plant-pollinator associations. In any case, it is worth noting that other local factors such as species relative abundances, surely explain part of the observed variability [@vazquez2007; @encinas2012; @bartomeus2016] that reproductive trade-offs do not.

To conclude, we provide a robust description of plant reproductive trade-offs using a large global dataset of plant traits. We identified the major reproductive strategies of flowering plants and how these strategies influence interactions with different floral visitor guilds. Although the explained variation that we found in the first two axes is lower than previous studies of vegetative traits [@diaz2016; @carmona2021] it is consistent with the largest and most recent study that has characterised plant life strategies with vegetative and reproductive traits [@salguero2016]. Future work needs to integrate the reproductive compromises that we have identified with vegetative and physiological trade-offs to create a more comprehensive spectrum of plant trait variation. Further, the varying level of phylogenetic signal among traits deserves further attention to understand evolutionary changes on mating and flower morphology in response to pollinators [@gervasi2017; @mackin2021]. Finally, including plant-pollinator networks from unrepresented areas of the world and a more complete description of plant reproductive trade-offs is essential for a better understanding of the global patterns in plant-pollinator interactions.

# References

<div id ="refs"></div>

# Acknowledgements

This study was supported by the European project SAFEGUARD (101003476 H2020-SFS-2019-2). We thank all researchers that made their data openly available and sent it upon request. We also thank Bryony Wilcox, Greg Bible, Mercedes Sanchez-Lanuza and David Ragel for their help with data collection. We also thank Jason Tylianakis for his comments on the manuscript before submission. Finally, JBL thanks the University of New England for the funding provided to carry out this work.

\singlespacing

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE, fig.height=1, fig.width=14}
setwd("~/R_Projects/Reproductive Traits")

library(kableExtra)

library(dplyr)
library(readxl) #read excel file (trait data)
library(dplyr) #data manipulation
library(visdat) #VISUALIZE MISSING DATA
library(naniar)
library(tidyverse)
########################################################################################################################################################
#1) READ TRAIT DATA
########################################################################################################################################################
#load data
trait_data <- read_excel("Data/Trait_data_raw/Trait_data_final.xlsx",na = "NA")
########################################################################################################################################################
#2) A) CLEAN DATA AND B) RENAME LEVELS FOR IMPUTATION 
########################################################################################################################################################
#####
#A) CLEAN DATA
#####
#select just filled rows
trait_data_1 <- trait_data[1:1712,] #It may be a more elegant way but this does the job, 1712 rows of data

#filter data, select species with flower level info and capitulum
trait_filtered <- filter(trait_data_1, Info_level == "flower" |  Info_level == "capitulum")
#remove NA's and duplicated species|Some spp are repeated once the spp names are standardize with taxize
trait_filtered$Species_all[trait_filtered$Species_all=="NA"]<-NA 
trait_filtered_1 <- trait_filtered[!is.na(trait_filtered$Species_all),]
trait_filtered_2 <- trait_filtered_1[!duplicated(trait_filtered_1$Species_all),]

#Select columns to work with
t <- trait_filtered_2[c("Species_geonet","Order_all","Family_all","Genus_all","Species_all","Breeding_system","Compatibility_system",
                        "Autonomous_selfing_level","Autonomous_selfing_level_fruit_set", "Flower_morphology", "Flower_symmetry", 
                        "Flowers_per_plant","Floral_unit_width", "Corolla_diameter_mean", "Corolla_length_mean","Style_length", "Ovule_number", 
                        "life_form", "lifespan","Plant_height_mean_m","Nectar_presence_absence","Nectar_ul","Nectar_mg","Nectar_concentration","Pollen_per_flower")]

>>>>

---
title: "Untitled"
output: html_document
---
# Regression trees
